Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-03-15 rmarkdown/paper.Rnw
new commit test

% \thanks{These two authors contributed equally.}

@@

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\cite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments that the protein under study interacts with using random
fragmentation of DNA and a protein-specific antibody. Then,
exonuclease is introduced to trim the $5\prime$ end of each DNA
fragment to a fixed distance from the bound protein compared to
ChIP-Seq. This step is unique to ChIP-exo and could potentially
provide significantly higher spatial resolution compared to
ChIP-Seq. Finally, high throughput sequencing of a small region (36 to
100 bp) at the $5\prime$ end of each fragment generates millions of
reads.

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol seem to be less complex than the
libraries generated by ChIP-Seq (Mahony et al., 2015,
\cite{exo_review}), i.e. the possible number of positions to which the
reads can be aligned has been reduced due to the exonuclease
digestion. Second, although there are roughly the same amount of read
in both strands, locally there may be more reads in one strand than in
the other. To address these challenges, we suggest a collection of
quality control visualizations to interrogate these biases in a
ChIP-exo experiment and globally asses the enrichment and library
complexity of a ChIP-exo sample. We gathered ChIP-exo data from
diverse organisms: CTCF factor in human \cite{exo1}; ER factor in
human and FoxA1 factor in mouse (Serandour et al., 2013
\cite{exoillumina}); Glucocorticoid receptor (GR) in IMR90, K562 and
U20S cell lines (Starick et al., 2015 \cite{starick15}).Furthermore,
we also generated ChIP-exo and ChIP-Seq data for $\sig$ factor in
\emph{Escherichia Coli} (\emph{E. Coli}) under aerobic ($ + O_2$)
condition, and treated by rifampicin by 0 and 20 minutes (courtesy of
Professor Robert Landick's lab).
%% note to self, we may add more experiments here

Most of current ChIP-Seq quality control (QC) guidelines (Landt et
al., \cite{encode_qc}) may not be applicable on ChIP-exo, additionally
to our knowledge there are not established QC pipelines for
ChIP-exo. Previous ChIP-exo analyses used ChIP-Seq samples to compare
the resolution between experiments (\cite{exo1}, \cite{exo2},
\cite{exoillumina}); Carroll et al., 2014 \cite{carroll.qc} studied
the use of the Strand-Cross Correlation (SCC) (Kharchenko et al., 2008
\cite{strandcc}) and showed that by filtering blacklisted regions the
estimation of the SCC is improved. However, using the SCC may not be
helpful since the peaks that are attained at the read and fragment
lengths are confused in a typical ChIP-exo SCC curve. Additionally
this method requires to know blacklisted regions in advance which may
not be available. In our pipeline we propose an out-the-shelf analysis
that contrasts the enrichment of the experiment against its library
complexity.

>>>>

@@

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\cite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments that the protein under study interacts with using random
fragmentation of DNA and a protein-specific antibody. Then,
exonuclease is introduced to trim the $5\prime$ end of each DNA
fragment to a fixed distance from the bound protein compared to
ChIP-Seq. This step is unique to ChIP-exo and could potentially
provide significantly higher spatial resolution compared to
ChIP-Seq. Finally, high throughput sequencing of a small region (36 to
100 bp) at the $5\prime$ end of each fragment generates millions of
reads. Figure \ref{fig:chip_diagram} illustrates the differences
between ChIP-exo, Single End (SE) ChIP-Seq and Paired End (PE)
ChIP-Seq: The $5\prime$ ends of a ChIP-exo experiment are located more
tightly around the binding proteins than in a ChIP-Seq experiment; in
a PE ChIP-Seq experiment both ends are observed while in a SE ChIP-Seq
experiment only the $5\prime$ end.

% Reffers to... First, DNA libraries... I think that these might be
% too much details for the introduction. How about discussing them in
% the results when we investigate the related

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol seem to be less complex than the
libraries generated by ChIP-Seq (Mahony et al., 2015,
\cite{exo_review}), i.e. the possible number of positions to which the
reads can be aligned has been reduced due to the exonuclease
digestion. Second, although there are roughly the same amount of read
in both strands, locally there may be more reads in one strand than in
the other.

To address these challenges, we suggest a collection of
quality control visualizations to interrogate these biases in a
ChIP-exo experiment and globally asses the enrichment and library
complexity of a ChIP-exo sample. We gathered ChIP-exo data from
diverse organisms: CTCF factor in human \cite{exo1}; ER factor in
human and FoxA1 factor in mouse (Serandour et al., 2013
\cite{exoillumina}); Glucocorticoid receptor (GR) in IMR90, K562 and
U20S cell lines (Starick et al., 2015 \cite{starick15}).Furthermore,
we also generated ChIP-exo and ChIP-Seq data for $\sig$ factor in
\emph{Escherichia Coli} (\emph{E. Coli}) under aerobic ($ + O_2$)
condition, and treated by rifampicin by 0 and 20 minutes (courtesy of
Professor Robert Landick's lab).
%% note to self, we may add more experiments here

Most of current ChIP-Seq quality control (QC) guidelines (Landt et
al., \cite{encode_qc}) may not be applicable on ChIP-exo, additionally
to our knowledge there are not established QC pipelines for
ChIP-exo. Previous ChIP-exo analyses used ChIP-Seq samples to compare
the resolution between experiments; Carroll et al., 2014
\cite{carroll.qc} studied the use of the Strand-Cross Correlation
(SCC) (Kharchenko et al., 2008 \cite{strandcc}) and showed that by
filtering blacklisted regions the estimation of the SCC is
improved. However, using the SCC may not be helpful since the peaks
that are attained at the read and fragment lengths are confused in a
typical ChIP-exo SCC curve. Additionally this method requires to know
blacklisted regions in advance which may not be available. In our
pipeline we propose an out-the-shelf analysis that contrasts the
enrichment of the experiment against its library complexity.
